{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-03-06 Docs/manuscript.Rnw
Results are edited, but we need to add model fits inline in the text and update tables

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{November 2022}
\fancyhead[RO]{Manuscript}
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models. We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain. We checked for model fit by ...
%emw2Mar -- Check: Did you use 4 chains? And add info. It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
%emw2Mar -- Did you use default priors? If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)? Ideally they don't and then we can add 'increasing priors three-fold did not change results.'

\section{Results}
% START HERE Lizzie ... 
\verb@Research questions recap@
\\
Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.

@@

%emw24Jan -- below is great opening!
Our final dataset captures seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019. Seven species had fall event information available. 
\\

\todo[inline]{add map here}
\todo[inline]{add a table documenting studies attributes}

\subsection{Provenance latitude and MAT do not affect spring event timing}
Overall, our models show that spring events are not related to provenance latitude or mean annual temperature in North America, and only weakly related in Europe, where springs events are slightly earlier where provenance mean annual temperature is lower (higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}).

@@

\subsubsection {Similar results across provenance latitude, absolute value of difference between provenance and garden latitude, and spherical distance}
Along with (a) provenance latitude, we also looked at how spring event timing is related to (b) the absolute value of difference between provenance and garden latitude , as well as (c) the spherical distance between the garden and each provenance. All three metrics depict little to no relationship between spring event timing and the geographical location of the provenance (Supplement Fig.\ref{supplement-figure:lat_distance}).

\newpage
\subsubsection {Climate overlap does not predict event dates much better than provenance latitude or MAT}

While comparing how similarity in climate relates to event dates, we observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms. Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
    \caption{Caption placeholder}
    \label{figure:overlap}
\end{figure}

\subsection{Stronger clines in fall events observed in North America than Europe}
%emw24Jan -- Nice! Also, results seem good, I can help trim text and collapse some sections.

@@

We find that fall events (budset, leaf senescence, leaf abcission) advance strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (higher, more northern latitudes). This relationship, however, is observed mostly in North America where fall events advance 4.2 days per degree we move north, or 6.4 days when the MAT decreases by 1 $^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance 0.5 days per degree we move north, or 0.6 days when the MAT decreases by 1 $^{\circ}$C(Fig. \ref{figure:continent}).

\subsection{Effects of MAT on spring events diverge across angiosperms and gymnosperms}

Effects of provenance latitude on both spring and fall events are similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS. 6.2 days)(Fig. \ref{figure:spp_type}).

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/spp_type_histogram.png}
    \caption{Caption placeholder}
    \label{figure:spp_type}
\end{figure}

>>>>

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{March 2023}
\fancyhead[RO]{Manuscript}
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used hierarchical Bayesian models. We regressed day of year (DOY) of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}), using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals, unless otherwise stated. 
%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves. 
%emw2Mar -- Did you use default priors? If so, can you figure out how to increase them threefold and see if answers change (they change run to run, but should not change dramatically)? Ideally they don't and then we can add 'increasing priors three-fold did not change results.'

\section{Results}
%emw6Mar -- below opening and questions are good, but not appropriate for results. Move up to be at end of introduction. 
\verb@Research questions recap@
\\
Despite a growing interest in predicting local adaptation across locations, no study has comprehensively examined clines for spring and fall events and what factors may underlie differences observed across studies.

@@

%emw24Jan -- below is great opening!
%emw6Mar -- results are in past tense generally
Our final dataset captured seven angiosperm and eight gymnosperm species from 19 common gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species also had fall event information available. 
\\

\todo[inline]{add map here}
\todo[inline]{add a table documenting studies attributes} %emw6Mar -- Tables should go in supplement. But we should add estimates to statements we make throughout. These should be added as mean +/- 90% unvertainty (posterior) intervals (UI) parenthetically using /Sexpr so they update automatically. 
Overall, spring events were not related to provenance latitude or mean annual temperature in North America (insert the mean slope here, followed by the 90\% intervals ), and only weakly related in Europe (insert slope and UI), where springs events are slightly earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude). %emw6Mar -- this text from below -- "advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower" is good, can you copy that phrasing here instead "where springs events are slightly earlier where provenance mean annual temperature is lower"? I think it would be something like 'delay slightly ...'

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and mean annual temperature, meaning fall events are earlier where provenance mean annual temperature is lower (i.e., higher, more northern latitudes). This relationship, however, was observed mostly in North America where fall events advance 4.2 days per degree (UI), or 6.4 days (UI) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance 0.5 days (UI) per degree we move north, or 0.6 days (UI) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}). %emw6Mar -- above numbers (e.g., 4.2 should be done with /Sexpr and include UI in parentheses)

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS. 6.2 days)(Fig. \ref{figure:spp_type}). %emw6Mar -- Again, need numbers using /Sexpr here.

We also tested whether climate effects on events may be stronger using metrics from seasonal windows closer to event dates, but results were not qualitatively different than using MAT. We observed very weak effects of climate overlap on spring events, nearly identical across angiosperms and gymnosperms. Fall events diverge as climate overlap declines for both angiosperms and gymnosperms, but slightly more strongly for gymnosperms (Fig.\ref{figure:overlap}).

%emw6Mar -- I think we can skip subheaders in the results (yay!) but I need to read it again once you had added all the numbers and intervals. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/spp_type_histogram.png}
    \caption{Caption placeholder}
    \label{figure:spp_type}
\end{figure}

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/climate_overlap.png}
    \caption{Caption placeholder}
    \label{figure:overlap}
\end{figure}
%emw6Mar -- Climate overlap figures can go in supp.

%emw6Mar -- I could combine the current fig 3-4. Just be sure to label each panel.

%emw24Jan -- Nice! Also, results seem good, I can help trim text and collapse some sections.

@@
